---
title: "Problem Set 3"
author: "Mariah Boudreau and Ben Camber"
date: "9/20/2019"
output: pdf_document
---
```{r, echo=FALSE}
load("COMADRE_v.3.0.0.RData")
```

## 1) What species (and location if available) did you choose? Why did you choose the species?

We choose the Eastern Hellbender, the largest salamander. They live in the Appalachian Mountain range. These appear to be a very intersting species that I have never learned about before. The paper for this model comes from the paper at this link https://www.sciencedirect.com/science/article/pii/S161713811300054X.


```{r, echo=FALSE}
comadre$mat[203][[1]]$matA

# comadre$matrixClass[203]

# comadre$metadata$Journal[203][[1]]

# comadre$metadata$Authors[203][[1]]

```

The matrix describes the stages of Eggs/Larvae, Sub-adults, and Adults, respectively.

## 2) What question do you want to answer about this population (e.g. population status, best management strategies)?

We can ask which stage is the most sensitive to intervention, then we could aid the population of this Eastern Hellbender. 

## 3) Calculate eigenvalue, stable age distribution, elasticity, and sensitivity. What does this tell you about the population?

```{r}
EHMat = comadre$mat[203][[1]]$matA
EHMat
eigen(EHMat)
standardEigenvec <- eigen(EHMat)$vectors[,1]/(eigen(EHMat)$vectors[1,1]+eigen(EHMat)$vectors[2,1]+eigen(EHMat)$vectors[3,1])

sensitivity <- function(A) {
  d <- eigen(A)$values   # eigen values
  w <- eigen(A)$vectors  # right eigen vectors
  v <- Conj(solve(w))    # complex conjugate of left eigen vectors
  # output of eigenvalues is decreasingly sorted.
  v[1,] %*% t(w[,1])
}
S <- sensitivity(EHMat)
S

L1 <- Re(eigen(EHMat)$values[1])
elas <- (EHMat/L1) * S
E <- round(elas, 3)
E

```

The survival of embryo/larval is 0.10, the probability of remaining a subadult ($P_i$) is 0.711, and the probability of juvenile becoming adult is 0.038.  The adult survival rate is 0.80. 

Our dominant eigenvalue was the value of `r eigen(EHMat)$values[1]`. This interprets to the Easten Hellbender's popultation growing at a rate of approximately 2.75%. The standardized eigenvector associated with this eigenvalue is `r standardEigenvec`. The interpretation for this standard eigenvector is the standard age distribution where there are about 84% of the population in the Egg/Laval stage, about 15% of the population in the Subadult/Juvenile stage, and about 1% in the Adult stage.

The sensitivity for this matrix is given above in the variable S. The elasticity for this matrix is given above in the variable E.

The sensitivity matrix shows that changing probability of a juvenile becoming an adult will increase the rate of growth the most out of all the parameters. The next most sensitive parameter is the survival rate of eggs/laval. The 31.86 in row 3 and column 1 would change the rate of growth the most, however, there is no way for eggs/larvae to transition directly to adulthood. The elasticity matrix shows that the proportional sensitivity is highest in the adult survival rate, and the probability of a subadult remaining a subadult. 

## 4) Using the calculations in part (c), or additional calculations, address the question you proposed in part (b).

The best way to manage Eastern Hellbender populations seems to be ensuring that juveniles survive to adulthood and that adults persist for a long time. Fecundity and egg survival parameters are insensitive and inelastic, which is consistent with what we know about organisms with an "r selected" reproduction strategy. Interestingly, the model was most sensitive at the transition from egg/larva to adult, and while that can't happen in nature, we could (and do) facilitate this by catching wild EH larvae and allowing them to mature in captivity, then releasing adults into the wild to breed.


## Easton comments

- Overall, really nice job on this assignment. You receive full points. The calculations were done correctly and you did a nice job at interpreting the math into biology. 
- The in-line R code for question 3 looks really nice as well
- To make pretty matrices (and tables) in R you can also try the xtable or kable packages
- I modified your first code chunk to call data not from the downloads folder. It is important to use "relative paths" when collaborating with others. 