---
title: "Problem Set 3"
author: "Mariah Boudreau"
date: "9/20/2019"
output: pdf_document
---
```{r, echo=FALSE}
load("~/Downloads/COMADRE_v.3.0.0.RData")
```

## 1) What species (and location if available) did you choose? Why did you choose the species?

We choose the Eastern Hellbender, the largest salamander. They live in the Appalachian Mountain range. These appear to be a very intersting species that I have never learned about before. The paper for this model comes from the paper at this link https://www.sciencedirect.com/science/article/pii/S161713811300054X.


```{r}
comadre$mat[203][[1]]

comadre$matrixClass[203]

comadre$metadata$Journal[203][[1]]

comadre$metadata$Authors[203][[1]]

```

The matrix describes the stages of Eggs/Larvae, Sub-adults, and Adults.

## 2) What question do you want to answer about this population (e.g. population status, best management strategies)?

We can ask which stage is the most sensitive to intervention, then we could aid the population of this Eastern Hellbender. 

## 3) Calculate eigenvalue, stable age distribution, elasticity, and sensitivity. What does this tell you about the population?

```{r}
EHMat = comadre$mat[203][[1]]$matA
EHMat 
eigen(EHMat)


```


