---
title: "USGS NAS"
author: "Mariah Boudreau, Kerria Burns, Amanda Jones"
date: "12/5/2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---
#Introduction

Zebra mussels belong to the phylum Mollusca, the group bivalves (unique among them--the only species that attaches to a hard substrate), which are native to the Black, Caspian, and Azov seas (USGS NAS). They were first observed in areas nearby Indiana, Michigan, and Ohio in 1986 (USGS NAS). Zebra mussels likely originated in Europe, establishing a population in Lakes St. Clair and Erie when ships dispersed ballast water (Hebert et al. 1989). The influence of bivalves on ecosystems includes the following: removal of particles from the water column, reducing populations of consumers depending on these particles as food source, increasing populations that use bivalves or their waste products, and making available particles previously used by phytoplankton (Strayer et al. 1999). As a result, their introduction into the United States as a result of commercial trade has had cascading effects on ecosystems, affecting multiple trophic levels and altering the abiotic conditions of the water bodies in which they reside.
	
The outcome of invasions depends in part on the tolerance of environmental conditions in the area of introduction. Introduced environments with conditions similar to those found in the native range may be more conducive to invasive establishment (Baker and Stebbins, 1965). Additionally, reproductive and physiological characteristics of the invading organism play a role in whether populations become established in the new range. In the case of zebra mussels, free swimming larvae and high fecundity (eggs per female) have been implicated in its proliferation success (Hebert et al. 1989). That zebra mussels are amenable to a wide range of habitats with a flexible reproductive system has aided their spread and establishment in North America where they were introduced via the release of ballast water from ships (Nichols 1996).

Physiological constraints of organisms determine the optimal range in which reproduction can occur, an important factor in the spread and establishment of an invasive species (or any species, for that matter). One such constraint is thermal tolerance, and temperature was shown to affect the timing of gametogenesis (Wacker & Elert 2003). The two experimental groups were raised at different depths to generate the temperature difference, and variation in environmental quality in the surrounding region may explain the differences in egg mass released at the two different depths (Wacker & Elert 2003). In addition, metabolic rates increase with temperature, so that differences in food availability at the two depths could further influence reproductive investment. Without respective to temperature differences, availability of polyunsaturated fatty acids and food quality in general resulted in changes to reproductive investment (Wacker & Elert 2003). An additional study confirms a threshold of 12 C for the onset of spawning and further identifies two spawning cohorts as the season proceeds from May to August (Borcherding 1991). Gametes are released into the water column over a period of 6 to 8 weeks, where they are fertilized and develop—30,000 to 40,000 eggs may be released by one female, though the actual number may be closer to 1.5 million. Egg release corresponds with temperature, beginning at 12C and peaking at 22C. As a result, juvenile proliferation should track these temperature changes (Hebert et al. 1988). High temperatures alongside low food availability results in reduction in the size of the gonads (Borcherding 1991).

Dreissena tends to colonize structures below 1.2 m. Zebra mussels may spread to larger structures through water intake pipes. Zebra mussels tend to increase water clarity through digestion of suspended sediments—for this reason zebra mussels have been intentionally stocked in lakes outside North America. Food deprived mussels fed indiscriminately on particles of all sizes, but satiated individuals felt only on those in a much smaller size range (MacIsaac 1996).

Ambient temperature, seston concentration, and mussel size frequency are three factors that influence Dreissena filtering impact. Maximal filtering rate has been hypothesized to be 5 and 20 C, declining outside of this range. Ingestion rate may also be temperature dependent; however, the results of experiments investigating the effect of temperature are variable. It may still be an important performance factor, and impose a range limitation in southern states. Depending on size, zooplankton may succumb to or evade ingestion by zebra mussels—smaller individuals cannot escape the inflow current and are not rejected, whereas larger individuals may dodge the current or be expelled as a result of irritating the feeding apparatus (MacIsaac 1996).






```{r, echo=FALSE, include=FALSE}
dat <- read.csv("NAS-Data-Download.csv")
library(dplyr)
library(ggplot2)

```


```{r, echo = FALSE, eval=FALSE}

# Checking to unique HUC ID Numbers
as.data.frame(table(dat$HUC.8.Number))
# There are 350 unique HUC ID numbers
# That means there are 350 different locations for this data
```


```{r, echo = FALSE}
# This chunk will be to filter the unnecessary parts of the data
fil_dat <- select(dat, Specimen.Number, Species.ID, Group, Family, Scientific.Name, 
                  Common.Name, Latitude, Longitude, Source, Accuracy, Drainage.Name,
                  HUC.8.Number, Year, Month, Day, Status)

```

#Methods 

##Initial Visualization
```{r, echo = FALSE, warning=FALSE}
# Just looking at some different ways to visualize and understand the data
ggplot(fil_dat, aes(x=dat$HUC.8.Number, y=dat$State)) + geom_point(aes(col=dat$Year), size=3) +labs(title="HUC Number vs. State", subtitle="From USGS Zebra Mussel Dataset", y="State", x="HUC Number", col = "Years") 
```

**Figure 1**
This graph is one way in which the data set from the USGS zebra mussel data can be visualized. The horizontal axis represents the HUC Number and it is plotted again the state in which that HUC resides. The color of the point exhibits the time when the data was collected. There appears to be an even spread of years throughout the plot given by the various color gradients. 

```{r,echo=FALSE, warning=FALSE}
# ggplot(fil_dat, aes(x=dat$HUC.8.Number, y=dat$Year)) + geom_point(aes(col=dat$State), size=3) +labs(title="HUC Number vs. Year", subtitle="From USGS Zebra Mussel Dataset", y="Year", x="HUC Number")


counts <- table(fil_dat$HUC.8.Number)
barplot(counts, main="HUC Distribution", 
   xlab="HUC", ylab = "Count")

# Look into the other large points from the HUC Distribution

# max(counts) #684 This occurs in Oneida Lake in NY
# which(table(fil_dat$HUC.8.Number) == 684) #4140202
# 
# head(fil_dat[fil_dat$HUC.8.Number == 414020200,])
# 
# 
# table(fil_dat$HUC.8.Number)


```

**Figure 2**
Number of zebra mussel occurrences per Hydraulic Unit Code.


```{r, echo=FALSE, warning=FALSE}
#dat <- read.csv("C:/Users/kmb057/Documents/FundamentalsQuantReasoning/QuantReasoning/NAS-Data-Download.csv")

#remove 0,0 coordinate row
dat <- dat[-5810,]

library(ggplot2)
library(maps)
library(gganimate)
library(dplyr)

#map of the 50 states
US <- map_data("state") 

states <- ggplot(data = US) + 
  geom_polygon(aes(x = long, y = lat, group = group), color = "white") + 
  coord_fixed(1.3) +
  guides(fill=FALSE)

#show change through time from earliest to latest
dat$Year <- sort(dat$Year,decreasing=F)

map1 <- states + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=dat,alpha=0.2)+transition_states(dat$Year)

#population in 2019
latest <- dat[dat$Year=="2019",]
map2 <- states + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=latest,alpha=0.2)

#animation of only those established or collected after 2000
established_collected_post_2000 <- filter(dat[dat$Year>=2000,]) %>%
  filter(Status=="established"|Status=="collected")

established_collected_post_2000$Year <- sort(established_collected_post_2000$Year,decreasing=F)

map3 <- states + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=established_collected_post_2000,alpha=0.2)+transition_states(established_collected_post_2000$Year)+shadow_mark()
```


```{r,echo=FALSE, warning = FALSE}


Summarize_sightings_by_year_month <- dat %>% count(Year, Month,State)
ggplot(Summarize_sightings_by_year_month, aes(fill=Summarize_sightings_by_year_month$State, y=Summarize_sightings_by_year_month$n, x=Summarize_sightings_by_year_month$Year)) +labs(x = "Year",y="Occurrences",colour="State") + geom_bar(position="dodge", stat="identity")+theme(legend.position = "none")
```


**Figure 3**
This graph demonstrates the number of occurrences by state, month, and year. The spread from state to state is observable.
 
```{r,echo=FALSE, warning = FALSE}
#filter data for only Great Lakes lat and long
BorderingGreatLakes <- c("New York","Pennsylvania","Ohio","Indiana","Michigan","Illinois","Wisconsin","Minnesota")
PlottingBorder <- map_data("state", region=BorderingGreatLakes)

Borders <- ggplot(data = PlottingBorder) + 
  geom_polygon(aes(x = long, y = lat, group = group), color = "white") + 
  coord_fixed(1.3) +
  guides(fill=FALSE)+scale_y_continuous(breaks=seq(from=30,to=50,by=1))+scale_x_continuous(breaks=seq(from=-100,to=-70,by=1))


Superior <- dat %>% filter(Latitude>=46 & Latitude<=50) %>% filter(Longitude<=-84.5 & Longitude>=-92)
map4 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Superior,alpha=0.2)

Michigan <- dat %>% filter(Latitude>=42.5 & Latitude<=46) %>% filter(Longitude<=-84.5 & Longitude>=-88)
map5 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Michigan,alpha=0.2)

Huron <- dat %>% filter(Latitude>=43 & Latitude<=46) %>% filter(Longitude<=-81.5 & Longitude>=-84)
map6 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Huron,alpha=0.2)

Erie <- dat %>% filter(Latitude>=41.5 & Latitude<=43) %>% filter(Longitude<=-79 & Longitude>=-84)
map7 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Erie,alpha=0.2)

Ontario <- dat %>% filter(Latitude>=43.25 & Latitude<=44) %>% filter(Longitude<=-78 & Longitude>=-80)
map8 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Ontario,alpha=0.2)
  
 #all great Lakes plotted
 GreatLakes <- Borders + 
  geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Superior,alpha=0.2)+
  geom_point(aes(x=Longitude,y=Latitude,color="blue"),data=Michigan,alpha=0.2)+
  geom_point(aes(x=Longitude,y=Latitude,color="green"),data=Huron,alpha=0.2)+
  geom_point(aes(x=Longitude,y=Latitude,color="yellow"),data=Erie,alpha=0.2)+
  geom_point(aes(x=Longitude,y=Latitude,color="purple"),data=Ontario,alpha=0.2)
 
 GreatLakes
```


**Figure 4**
This map shows the occurrences for only the Great Lakes region. Notably, Lake Michigan has the most, and Lake Superior has the least.
```{r,echo=FALSE}
#number of specimens found in each lake`
sup_occ <- unique(Superior$Specimen.Number)
mich_occ <- unique(Michigan$Specimen.Number)
huron_occ <- unique(Huron$Specimen.Number)
erie_occ <- unique(Erie$Specimen.Number)
ont_occ <- unique(Ontario$Specimen.Number)
zm_occ <- c(sup_occ, mich_occ, huron_occ, erie_occ, ont_occ)

SupOcc <- nrow(table(sup_occ))
MichOcc <- nrow(table(mich_occ))
HurOcc <- nrow(table(huron_occ))
EriOcc <- nrow(table(erie_occ))
OntOcc <- nrow(table(ont_occ))
zm_occ <- c(SupOcc, MichOcc, HurOcc, EriOcc, OntOcc)
```
#Results
One occurrence indicates one collection effort--anywhere from one to thousands of zebra mussels could be included. Lake Michigan has the highest number of occurrences at 610, while Lake Superior has the least, at 36.
```{r, echo = FALSE, warning = FALSE}

CountYear <- lm(n~Year, Summarize_sightings_by_year_month)
#summary(CountYear)
ggplot(Summarize_sightings_by_year_month, aes(x=Summarize_sightings_by_year_month$Year, y=Summarize_sightings_by_year_month$n)) + geom_point() + geom_smooth(method = "lm") + labs(title="Occurrences By Year", y="Count",x="Year")

```


**Figure 5**
There is no relationship between number of occurrences and year. Number of sightings does not increase or decrease as time goes on.

##Linear Regression of Size of Lake and HUC Counts
```{r SizeLR, echo=FALSE, warning=FALSE}
#Want to use this chunk to see if there is a correlation between number of occurances in each lake and their size. Linear Regression Analysis. If there isn't a positive correlation then would it be more around tourist areas? or more populated with boating areas. 
 
 # https://www.britannica.com/place/Great-Lakes
 
 #Superior, Michigan, Huron, Erie, Ontario sizes in square miles
 size_of_lake <- c(31700, 22300, 23000, 9910, 7340)
 
 
 sup_counts <-nrow(table(Superior$HUC.8.Number))
 mich_counts <-nrow(table(Michigan$HUC.8.Number))
 huron_counts <-nrow(table(Huron$HUC.8.Number))
 erie_counts <- nrow(table(Erie$HUC.8.Number))
 ont_counts <- nrow(table(Ontario$HUC.8.Number))
 zm_HUC_occurances <-c(sup_counts, mich_counts, huron_counts, erie_counts, ont_counts)
 
 sizeDat <- data.frame(size_of_lake, zm_HUC_occurances)

 
 HUCmodelSize <- lm(zm_HUC_occurances~size_of_lake, sizeDat)
 #summary(HUCmodelSize)
 sizeCoef <- coef(HUCmodelSize)[2]
 

 
 
 ggplot(sizeDat, aes(x=size_of_lake, y=zm_HUC_occurances)) + geom_point() + geom_smooth(method = "lm") + labs(title="Size of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset", y="HUC Occurrences", x="Size of Lake") 
 
 # There is a confidence band around the regression line as well. 

#number of specimens found in each lake`
sup_occ <- unique(Superior$Specimen.Number)
mich_occ <- unique(Michigan$Specimen.Number)
huron_occ <- unique(Huron$Specimen.Number)
erie_occ <- unique(Erie$Specimen.Number)
ont_occ <- unique(Ontario$Specimen.Number)
zm_occ <- c(sup_occ, mich_occ, huron_occ, erie_occ, ont_occ)

SupOcc <- nrow(table(sup_occ))
MichOcc <- nrow(table(mich_occ))
HurOcc <- nrow(table(huron_occ))
EriOcc <- nrow(table(erie_occ))
OntOcc <- nrow(table(ont_occ))
zm_occ <- c(SupOcc, MichOcc, HurOcc, EriOcc, OntOcc)
 
```


**Figure 6**
A linear regression analysis was conducted after collecting data on the size of the lake and filtering out the data to only look at the HUC zebra mussel occurrences in the areas of the Great Lakes. The reason for this analysis is to understand if there is a relationship between the size of the lake in water volume, and the number of occurrences on certain HUCs. 

Once the analysis was conducted, the p-value associated with the coefficient is not a significant value, and therefore not statistically significant. From the model, its states that a unit increase in the size of the lake leads to a `r sizeCoef` increase in the HUC Occurrences. Therefore, there is not a significant relationship between the HUC Occurrences, and the size of each Great Lake. The plot also depicts the trend line with the slope of `r sizeCoef` and there are confidences bands around the trend line also. This is a small data set with only five data points though, given there are only five Great Lakes. 



```{r separateSizeCode, echo = FALSE, warning = FALSE}
# This chunk will change the years that the HUC data is being looked at in 8 year chunks
 SuperiorFirst8 <- dat %>% 
   filter(Latitude>=46 & Latitude<=50) %>% 
   filter(Longitude<=-84.5 & Longitude>=-92) %>% 
   filter(Year <=1999 & Year >= 1991)
 
 SuperiorSecond8 <- dat %>% 
   filter(Latitude>=46 & Latitude<=50) %>% 
   filter(Longitude<=-84.5 & Longitude>=-92) %>% 
   filter(Year <= 2008 & Year >= 2000)

 SuperiorThird8 <- dat %>% 
   filter(Latitude>=46 & Latitude<=50) %>% 
   filter(Longitude<=-84.5 & Longitude>=-92) %>% 
   filter(Year <= 2015 & Year >= 2009)
 
MichiganFirst8 <- dat %>% 
  filter(Latitude>=42.5 & Latitude<=46) %>% 
  filter(Longitude<=-84.5 & Longitude>=-88) %>% 
   filter(Year <=1999 & Year >= 1991)

MichiganSecond8 <- dat %>% 
  filter(Latitude>=42.5 & Latitude<=46) %>% 
  filter(Longitude<=-84.5 & Longitude>=-88) %>% 
   filter(Year <=2008 & Year >= 2000)

MichiganThird8 <- dat %>% 
  filter(Latitude>=42.5 & Latitude<=46) %>% 
  filter(Longitude<=-84.5 & Longitude>=-88) %>% 
   filter(Year <=2015 & Year >= 2009)

HuronFirst8 <- dat %>% 
  filter(Latitude>=43 & Latitude<=46) %>% 
  filter(Longitude<=-81.5 & Longitude>=-84) %>% 
   filter(Year <=1999 & Year >= 1991)

HuronSecond8 <- dat %>% 
  filter(Latitude>=43 & Latitude<=46) %>% 
  filter(Longitude<=-81.5 & Longitude>=-84) %>% 
   filter(Year <=2008 & Year >= 2000)

HuronThird8 <- dat %>% 
  filter(Latitude>=43 & Latitude<=46) %>% 
  filter(Longitude<=-81.5 & Longitude>=-84) %>% 
   filter(Year <=2015 & Year >= 2009)

ErieFirst8 <- dat %>% 
  filter(Latitude>=41.5 & Latitude<=43) %>%
  filter(Longitude<=-79 & Longitude>=-84) %>% 
   filter(Year <=1999 & Year >= 1991)

ErieSecond8 <- dat %>% 
  filter(Latitude>=41.5 & Latitude<=43) %>%
  filter(Longitude<=-79 & Longitude>=-84) %>% 
   filter(Year <=2008 & Year >= 2000)

ErieThird8 <- dat %>% 
  filter(Latitude>=41.5 & Latitude<=43) %>%
  filter(Longitude<=-79 & Longitude>=-84) %>% 
   filter(Year <=2015 & Year >= 2009)

OntarioFirst8 <- dat %>% 
  filter(Latitude>=43.25 & Latitude<=44) %>% 
  filter(Longitude<=-78 & Longitude>=-80) %>% 
   filter(Year <=1999 & Year >= 1991)

OntarioSecond8 <- dat %>% 
  filter(Latitude>=43.25 & Latitude<=44) %>% 
  filter(Longitude<=-78 & Longitude>=-80) %>% 
   filter(Year <=2008 & Year >= 2000)

OntarioThird8 <- dat %>% 
  filter(Latitude>=43.25 & Latitude<=44) %>% 
  filter(Longitude<=-78 & Longitude>=-80) %>% 
   filter(Year <=2015 & Year >= 2009)

```


To conduct further analysis, that data used used the time span of the years 1991-2015 and the data was then split into three different different time period. The time periods are [1991-1999], [2000-2008], and [2009-2015]. Each time period has a linear regression analysis associated with it and the details for each will be described along with a graph of each trend line. 



```{r regressFirst8, echo = FALSE, warning=FALSE}
  ## This is the first 8 years of the data (1991-1999)

sup_counts_F8 <-nrow(table(SuperiorFirst8$HUC.8.Number))
 mich_counts_F8 <-nrow(table(MichiganFirst8$HUC.8.Number))
 huron_counts_F8<-nrow(table(HuronFirst8$HUC.8.Number))
 erie_counts_F8 <- nrow(table(ErieFirst8$HUC.8.Number))
 ont_counts_F8 <- nrow(table(OntarioFirst8$HUC.8.Number))
 zm_HUC_occurances_F8 <-c(sup_counts_F8, mich_counts_F8, huron_counts_F8, erie_counts_F8, ont_counts_F8)

 HUCmodelSizeF8 <- lm(zm_HUC_occurances_F8~size_of_lake, sizeDat)
 #summary(HUCmodelSizeF8)
 sizeCoefF8 <- coef(HUCmodelSizeF8)[2]


 
 sizeDatF8 <- data.frame(size_of_lake, zm_HUC_occurances_F8)

 
 ggplot(sizeDatF8, aes(x=size_of_lake, y=zm_HUC_occurances_F8)) + geom_point() + geom_smooth(method = "lm") + labs(title="Size of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset For First 8 Years", y="HUC Occurrences", x="Size of Lake") 
 
```


**Figure 7**
Number of HUC occurrences as a function of lake size for the first eight years.

The p-value associated with the coefficient from the data in the range of 1991-1999 is not a significant value, and therefore not statistically significant. From the model, its states that a unit increase in the size of the lake leads to a `r sizeCoefF8` increase in the HUC Occurrences between the years of 1991-1999. Therefore, there is not a significant relationship between the HUC Occurrences, and the size of each Great Lake during that time frame. The plot also depicts the trend line with the slope of `r sizeCoefF8` and there are confidences bands around the trend line also. 

```{r regressSecond8, echo = FALSE, warning=FALSE}
  ## This is the second 8 years of the data (2000-2008)

sup_counts_S8 <-nrow(table(SuperiorSecond8$HUC.8.Number))
 mich_counts_S8 <-nrow(table(MichiganSecond8$HUC.8.Number))
 huron_counts_S8<-nrow(table(HuronSecond8$HUC.8.Number))
 erie_counts_S8 <- nrow(table(ErieSecond8$HUC.8.Number))
 ont_counts_S8 <- nrow(table(OntarioSecond8$HUC.8.Number))
 zm_HUC_occurances_S8 <-c(sup_counts_S8, mich_counts_S8, huron_counts_S8, erie_counts_S8, ont_counts_S8)

 HUCmodelSizeS8 <- lm(zm_HUC_occurances_S8~size_of_lake, sizeDat)
# summary(HUCmodelSizeS8)
 sizeCoefS8 <- coef(HUCmodelSizeS8)[2]
 
 
 sizeDatS8 <- data.frame(size_of_lake, zm_HUC_occurances_S8)

 
 ggplot(sizeDatS8, aes(x=size_of_lake, y=zm_HUC_occurances_S8)) + geom_point() + geom_smooth(method = "lm") + labs(title="Size of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset For Second 8 Years", y="HUC Occurrences", x="Size of Lake") 
 
```


**Figure 8**
Number of HUC occurrences as a function of lake size for the second eight years.

The p-value associated with the coefficient from the data in the range of 2000-2008 is not a significant value, and therefore not statistically significant. From the model, its states that a unit increase in the size of the lake leads to a `r sizeCoefS8` increase in the HUC Occurrences between the years of 2000-2008. Therefore, there is not a significant relationship between the HUC Occurrences, and the size of each Great Lake during that time frame. The plot also depicts the trend line with the slope of `r sizeCoefS8` and there are confidences bands around the trend line also. This relationship is also negative compared the the first 8 years, which had a positive relationship. Even though neither of the slopes were significant, there could be a change occurring in those time periods that need to be explored. 

```{r regressThird8, echo = FALSE, warning = FALSE}
  ## This is the third 8 years of the data (2009-2015)

sup_counts_T8 <-nrow(table(SuperiorThird8$HUC.8.Number))
 mich_counts_T8 <-nrow(table(MichiganThird8$HUC.8.Number))
 huron_counts_T8<-nrow(table(HuronThird8$HUC.8.Number))
 erie_counts_T8 <- nrow(table(ErieThird8$HUC.8.Number))
 ont_counts_T8 <- nrow(table(OntarioThird8$HUC.8.Number))
 zm_HUC_occurances_T8 <-c(sup_counts_T8, mich_counts_T8, huron_counts_T8, erie_counts_T8, ont_counts_T8)

 HUCmodelSizeT8 <- lm(zm_HUC_occurances_T8~size_of_lake, sizeDat)

 # summary(HUCmodelSizeT8)
 sizeCoefT8 <- coef(HUCmodelSizeT8)[2]
 sizeDatT8 <- data.frame(size_of_lake, zm_HUC_occurances_T8)


 
 
 ggplot(sizeDatT8, aes(x=size_of_lake, y=zm_HUC_occurances_T8)) + geom_point() + geom_smooth(method = "lm") + labs(title="Size of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset For Third 8 Years", y="HUC Occurrences", x="Size of Lake") 
 
```


**Figure 9**
Number of HUC occurrences as a function of lake size for the third eight years.

The p-value associated with the coefficient from the data in the range of 2009-2015 is not a significant value, and therefore not statistically significant. From the model, its states that a unit increase in the size of the lake leads to a `r sizeCoefT8` increase in the HUC Occurrences between the years of 2009-2015. Therefore, there is not a significant relationship between the HUC Occurrences, and the size of each Great Lake during that time frame. The plot also depicts the trend line with the slope of `r sizeCoefT8` and there are confidences bands around the trend line also. This coefficient is negative and larger than the previous 8 year period. However, the overall regression analyis shows a positive non significant trend, which insinuates that the first 8 year period has a large affect on the regression analysis since it is the only positive trend out of the three time periods. 



```{r TempLR, echo = FALSE, warning = FALSE}
library(readxl)
lakes_92_18 <- read_excel("greatlakesavgtemp1992-2018.xlsx")
superiorAvgTemp <- sum(lakes_92_18$Superior)/366
michiganAvgTemp <- sum(lakes_92_18$Michigan)/366
huronAvgTemp <- sum(lakes_92_18$Huron)/366
erieAvgTemp <- sum(lakes_92_18$Erie)/366
ontarioAvgTemp <- sum(lakes_92_18$Ontario)/366

avgTemp <-c(superiorAvgTemp, michiganAvgTemp, huronAvgTemp,erieAvgTemp,ontarioAvgTemp)


 sup_counts <-nrow(table(Superior$HUC.8.Number))
 mich_counts <-nrow(table(Michigan$HUC.8.Number))
 huron_counts <-nrow(table(Huron$HUC.8.Number))
 erie_counts <- nrow(table(Erie$HUC.8.Number))
 ont_counts <- nrow(table(Ontario$HUC.8.Number))
 zm_HUC_occurances <-c(sup_counts, mich_counts, huron_counts, erie_counts,  
                       ont_counts)
 
 TempDat <- data.frame(avgTemp, zm_HUC_occurances)

 
 HUCmodelTemp <- lm(zm_HUC_occurances~avgTemp, TempDat)
 #summary(HUCmodelTemp)
 

 
 ggplot(sizeDatT8, aes(x=avgTemp, y=zm_HUC_occurances_T8)) + geom_point() + geom_smooth(method = "lm") + labs(title="Temperature of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset", y="HUC Occurrences", x="Temperature of Lake") 

```


**Figure 10**
This plot show the linear regression line that was fit to the data, along with the 95% confidence bands. Analysis to be done on this, not significant values here.

```{r, echo=FALSE, warning = FALSE}
 # Superior, Michigan, Huron, Erie, Ontario sizes in cubic miles
volume_of_lake <- c(2900, 1180, 850, 116, 393)


sup_counts <-nrow(table(Superior$HUC.8.Number))
mich_counts <-nrow(table(Michigan$HUC.8.Number))
huron_counts <-nrow(table(Huron$HUC.8.Number))
erie_counts <- nrow(table(Erie$HUC.8.Number))
ont_counts <- nrow(table(Ontario$HUC.8.Number))
zm_HUC_occurances <-c(sup_counts, mich_counts, huron_counts, erie_counts, ont_counts)

volumeDat <- data.frame(volume_of_lake, zm_HUC_occurances)


HUCmodelVolume <- lm(zm_HUC_occurances~volume_of_lake, volumeDat)
#summary(HUCmodelVolume)




ggplot(volumeDat, aes(x=volume_of_lake, y=zm_HUC_occurances)) + geom_point() + geom_smooth(method = "lm") + labs(title="Volume of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset", y="HUC Occurrences", x="Volume of Lake") 


```


**Figure 11**
This plot displays the linear regression line fitted for the HUC Occurrences and the volume of each lake. Analysis to be done at on this at another time. 

##Temperature Analysis
```{r Temp, echo = FALSE, warning=FALSE}
##  Temperature data
library(readxl)
library(dplyr)
library(ggplot2)
lakes_92_18 <- read_excel("greatlakesavgtemp1992-2018.xlsx")
every_lake <- read_excel("everylake1995-2018.xlsx")

#This graph shows the average temperatures for the lakes over the course of the year

plot(lakes_92_18$Day,lakes_92_18$Superior, col= "blue", xlim= c(1,365), ylim= c(0,30), 
     xlab= "Day of the Year", ylab= "Temperature (Celsius)", main= "Average Temp 1992-2018")
points(lakes_92_18$Day,lakes_92_18$Michigan, col= "orange")
points(lakes_92_18$Day,lakes_92_18$Huron, col="green" )
points(lakes_92_18$Day,lakes_92_18$Erie, col= "red")
points(lakes_92_18$Day,lakes_92_18$Ontario, col= "purple")
legend(0, 27, legend=c("Superior","Michigan","Huron","Erie","Ontario"), 
       col=c("blue","orange","green","red","purple"), lty=1, title= "Lake")


```


**Figure 12**
This graph depicts the average temperatures for the Great Lakes over the course of the year. Lake Superior is clearly colder than the rest of the Great Lakes, which is likely due to its higher latitude.


```{r Temp2, echo = FALSE, warning=FALSE, eval=FALSE}
#This shows the number of sightings per lake per year
library(stringr)
Sup_by_year <- dat %>%
  select(Locality,Year) %>%
  filter(str_detect(Locality,"Superior"))%>%
  count(Year)
plot(Sup_by_year, ylab= "Number of Sightings", main= "Sightings per Year", col= "blue", pch=19, ylim= c(0,100))
Mich_by_year <- dat %>%
  select(Locality,Year) %>%
  filter(str_detect(Locality,"Michigan"))%>%
  count(Year)
points(Mich_by_year, col= "orange", pch= 19)
Huron_by_year <- dat %>%
  select(Locality,Year) %>%
  filter(str_detect(Locality,"Huron"))%>%
  count(Year)
points(Huron_by_year, col= "green", pch=19)
Erie_by_Year <- dat %>%
  select(Locality,Year) %>%
  filter(str_detect(Locality,"Erie"))%>%
  count(Year)
points(Erie_by_Year,  col="red", pch=19)
Ont_by_Year <- dat %>%
  select(Locality,Year) %>%
  filter(str_detect(Locality,"Ontario"))%>%
  count(Year)
points(Ont_by_Year, col= "purple", pch= 19)
legend(2007, 100, legend=c("Superior","Michigan","Huron","Erie","Ontario"), 
       col=c("blue","orange","green","red","purple"), lty=1, title= "Lake")
```



```{r Temp3, echo = FALSE, warning=FALSE, eval=FALSE}
#Find high temp of each year of each lake
sup_high_temp <- every_lake%>%
  select(Year,Sup.)%>%
  group_by(Year)%>%
  filter(Sup.==max(Sup.))
plot(sup_high_temp, col= "blue", pch=19, ylim= c(10,30), ylab= "High Temp (Celsius)", 
     main= "High Temperature per Year")
Mich_high_temp <- every_lake%>%
  select(Year,Mich.)%>%
  group_by(Year)%>%
  filter(Mich.==max(Mich.))
points(Mich_high_temp, col="orange", pch=19)
Huron_high_temp <- every_lake%>%
  select(Year,Huron)%>%
  group_by(Year)%>%
  filter(Huron==max(Huron))
points(Huron_high_temp, col="green", pch=19)
Erie_high_temp <- every_lake%>%
  select(Year,Erie)%>%
  group_by(Year)%>%
  filter(Erie==max(Erie))
points(Erie_high_temp,col = "red", pch=19)
Ont_high_temp <-every_lake%>%
  select(Year,Ont.)%>%
  group_by(Year)%>%
  filter(Ont.==max(Ont.))
points(Ont_high_temp,col="purple", pch=19)

```


```{r Temp4, echo = FALSE, warning=FALSE}
#Correlation between high temp and number of sightings
lake_temp_counts <- read_excel("lakehightempandcounts.xlsx")

plot(lake_temp_counts$`Ont. Temp`,lake_temp_counts$`Ont. Count`, xlab= "High Temperature (Celsius)", 
     ylab= "Number of Sightings", main= "Temperature vs Number of Sightings", 
     col= "purple", pch=19, xlim= c(14,27), ylim= c(0,60))
points(lake_temp_counts$`Erie Temp.`,lake_temp_counts$`Erie Count`, col= "red", pch=19)
points(lake_temp_counts$`Huron Temp.`,lake_temp_counts$`Huron Count`, col= "green", pch=19)
points(lake_temp_counts$`Mich. Temp`, lake_temp_counts$`Mich. Count`, col= "orange", pch=19)
points(lake_temp_counts$`Sup. Temp`, lake_temp_counts$`Sup. Count`, col= "blue", pch=19)
abline(lm(lake_temp_counts$`Ont. Temp` ~ lake_temp_counts$`Ont. Count`), col= "purple")
abline(lm(lake_temp_counts$`Erie Temp.`~ lake_temp_counts$`Erie Count`), col= "red")
abline(lm(lake_temp_counts$`Huron Temp.`~lake_temp_counts$`Huron Count`), col= "green")
abline(lm(lake_temp_counts$`Mich. Temp`~lake_temp_counts$`Mich. Count`), col="orange")
abline(lm(lake_temp_counts$`Sup. Temp`~lake_temp_counts$`Sup. Count`), col= "blue")
legend(14, 60, legend=c("Superior","Michigan","Huron","Erie","Ontario"), 
       col=c("blue","orange","green","red","purple"), lty=1, title= "Lake")

#Gives R squared, p values, other statistics
#summary(lm(lake_temp_counts$`Ont. Temp` ~ lake_temp_counts$`Ont. Count`))
#summary(lm(lake_temp_counts$`Erie Temp.`~ lake_temp_counts$`Erie Count`))
#summary(lm(lake_temp_counts$`Huron Temp.`~lake_temp_counts$`Huron Count`))      
#summary(lm(lake_temp_counts$`Mich. Temp`~lake_temp_counts$`Mich. Count`))
#summary(lm(lake_temp_counts$`Sup. Temp`~lake_temp_counts$`Sup. Count`))

```


**Figure 13** 
The number of sightings per lake per year was compared with the high temperature of each lake for the years 1995-2018. The temperature data began with 1995, so no analyses could be performed on the sighting data prior to 1995. There are no significant correlations between the high temperature of the lakes and the number of sightings that occurred during any given year. Therefore, it can be concluded that temperature is not a driving factor of zebra mussel spread or a predictor for number of sightings in a given year. Further analysis may include temperature data for every lake throughout the entire year rather than just the highest recorded temperature for each year. The number of sightings is also not necessarily equivalent to the number of zebra mussels spotted.  

```{r Bias, echo=FALSE, warning=FALSE}
#Looking at sampling efforts per month

library(stringr)

Sup_by_month <- dat %>%
  select(Locality,Month) %>%
  filter(str_detect(Locality,"Superior"))%>%
  count(Month)
plot(Sup_by_month, ylab= "Number of Sightings", main= "Sightings per Month", col= "blue", pch=19, ylim= c(0,100))

Mich_by_month <- dat %>%
  select(Locality,Month) %>%
  filter(str_detect(Locality,"Michigan"))%>%
  count(Month)
points(Mich_by_month, col= "orange", pch= 19)

Huron_by_month <- dat %>%
  select(Locality,Month) %>%
  filter(str_detect(Locality,"Huron"))%>%
  count(Month)
points(Huron_by_month, col= "green", pch=19)

Erie_by_month <- dat %>%
  select(Locality,Month) %>%
  filter(str_detect(Locality,"Erie"))%>%
  count(Month)
points(Erie_by_month,  col="red", pch=19)

Ont_by_month <- dat %>%
  select(Locality,Month) %>%
  filter(str_detect(Locality,"Ontario"))%>%
  count(Month)
points(Ont_by_month, col= "purple", pch= 19)

legend(1, 100, legend=c("Superior","Michigan","Huron","Erie","Ontario"), 
       col=c("blue","orange","green","red","purple"), lty=1, title= "Lake")

```
**Figure 14**
There is a clear increase in sampling during the summer months (~July-October) across all five lakes.  


#Discussion
This analysis was mostly a descriptive visualization--where, and when, are zebra mussels found? Additional linear models demonstrate no significant relationship of zebra mussel occurrences to the size (area or volume) of the lake, or the time period. There are more sightings during the summer months (approx. July-October). This is likely due to the fact that most sampling is done during the summer when the lakes are not frozen. Therefore, there is a significant bias for what time of year there are the most sightings of zerba mussels. In general, zebra mussel sightings do not appear correlated with either volume or surface area of lake, the timeframe (year) collected, or temperature. The only correlation is between the time of year and occurrences, and this is likely due to logistical sampling based limitations. Further analyses could be conducted on the relationship of zebra mussel occurrences to latitude, or a comparison with biotic characteristics such as the presence of another invasive species, quagga mussels. One question in this regard could be how long after introduction of quagga mussels do zebra mussels persist before being outcompeted, since quagga mussels exist at much higher densities than zebra mussels. Projections under different temperature regimes based on global warming could also be carried out, though the analyses here demonstrate that temperatures found within this temperate biome is within the range tolerated by zebra mussels. However, zebra mussels may be affected by high temperatures in the southern United States, so predicted increasing temperatures may have an impact on populations in the future. Another area of analysis to conduct would be looking at boat travel and surveying lock use around the Great Lakes to see how this type of data relates to the zebra mussel sighting data. From the background information on zebra mussels, this type of analysis could confirm how the number of sightings has increased over the years depending on how much influx of boat travel in the Great Lakes there is. 


# Easton comments

In general, you did a nice job of finding a dataset, learning about the system, and running some analyses. Most of my comments below have to deal with the presentation of the results and organization of the text. 

Things to do for final draft:

- In general the figures look nice, but they can be cleaned up with little things like the axes titles
- Reorder the text at the beginning so it reads with a traditional introduction, methods, and results
- add appropriate figure captions
- include formatting to make R markdown output look nicer (e.g. section headers, remove chunk messages)
- Have a short paragraph that summarizes and puts your results in context


##References
Hebert, P. D. N., Mackie, G. L., & Muncaster, B. W. (1989). Ecological and Genetic Studies on Dreissena polymorpha ( Pallas ): A New Mollusc in the Great Lakes. Candian Journal of Fisheries and Aquatic Sciences. https://doi.org/10.1139/f89-202

Macisaac, H. J. (2014). Potential Abiotic and Biotic Impacts of Zebra Mussels on the Inland Waters of North America Potential Abiotic and Biotic Impacts of Zebra Mussels on the, (March). https://doi.org/10.1093/icb/36.3.287

Mackie, G. L. (2016). Biology of the exotic zebra mussel , Dreissena polymporph , in relation to native bivalves and its potential impact in Lake St . Clair, (July 1991). https://doi.org/10.1007/BF00024759

Nichols, S. J. (1996). Variations in the Reproductive Cycle of Dreissena Polymorpha in Europe, Russia, and North America. American Zoologist, 36, 311–325.

Strayer, D. L., Caraco, N. F., Cole, J. J., Findlay, S., & Pace, M. L. (1999). Transformation of Freshwater A case study of zebra mussels in the Hudson River. BioScience, 49(1).

Wacker, E. A. (2003). Food quality controls reproduction of the zebra mussel ( Dreissena polymorpha ). Oecologia, 135, 332–338. https://doi.org/10.1007/s00442-003-1208-5
>>>>>>> e30d40a615b2019c8289be6a267060cf2635fdd2
