---
title: "USGS NAS"
author: "Mariah Boudreau, Kerria Burns, Amanda Jones"
date: "12/5/2019"
output:
  html_document:
    df_print: paged
---
The source is the United States Geological Survey (USGS) Nonindigenous Aquatic 
Species (NAS) data set. It is a collection of zebra mussel sightings across the United States. 
Zebra mussels are an incredibly damaging invasive species in the continental US. They have 
spread drastically over the past several decades, beginning in the Great Lakes.  Included is a 
point map that represents the number of individuals by the color and size of the circle 
superimposed on a map of the United States. Also included is an Excel spreadsheet including 
species identification, latitude and longitude, phase of establishment, and date of sighting, among
other information.
 
Figure 1
Zebra mussel sightings represented by size and color of circle – darker and larger circle correspond to a greater number of individuals. Source: USGS NAS
The map includes Hydraulic Unit Codes (HUC), a system derived by USGS to delimit subregions within the country. 

An inquiry then could be to standardize each HUC by area to find the number of zebra mussels per unit area to determine at which locations the highest density occurs. Similarly, bins of degrees of latitude could be derived (to reduce the step of standardization and bypass the use of HUC altogether, a possible benefit of which would be that measurements of latitude might yield more climatically relevant data rather than political/economic divisions), to again determine the density of zebra mussels per unit (as yet to be defined). This could be combined with climatic information to determine which climate conditions favor the establishment and invasion of zebra mussels to the most severe extent. A time series could be constructed to see if there were certain time periods that showed more movement than others. The type of analysis could be done on a map similar to the one above to see the visual change as well. 

Zebra mussels belong to the phylum Mollusca, the group bivalves (unique among them--the only species that attaches to a hard substrate), which are native to the Black, Caspian, and Azov seas (USGS NAS). They were first observed in areas nearby Indiana, Michigan, and Ohio in 1986 (USGS NAS). Zebra mussels likely originated in Europe, establishing a population in Lakes St. Clair and Erie when ships dispersed ballast water (Hebert et al. 1989). The influence of bivalves on ecosystems includes the following: removal of particles from the water column, reducing populations of consumers depending on these particles as food source, increasing populations that use bivalves or their waste products, and making available particles previously used by phytoplankton (Strayer et al. 1999). As a result, their introduction into the United States as a result of commercial trade has had cascading effects on ecosystems, affecting multiple trophic levels and altering the abiotic conditions of the water bodies in which they reside.
	
The outcome of invasions depends in part on the tolerance of environmental conditions in the area of introduction. Introduced environments with conditions similar to those found in the native range may be more conducive to invasive establishment (Baker and Stebbins, 1965). Additionally, reproductive and physiological characteristics of the invading organism play a role in whether populations become established in the new range. In the case of zebra mussels, free swimming larvae and high fecundity (eggs per female) have been implicated in its proliferation success (Hebert et al. 1989). That zebra mussels are amenable to a wide range of habitats with a flexible reproductive system has aided their spread and establishment in North America where they were introduced via the release of ballast water from ships (Nichols 1996).

Physiological constraints of organisms determine the optimal range in which reproduction can occur, an important factor in the spread and establishment of an invasive species (or any species, for that matter). One such constraint is thermal tolerance, and temperature was shown to affect the timing of gametogenesis (Wacker & Elert 2003). The two experimental groups were raised at different depths to generate the temperature difference, and variation in environmental quality in the surrounding region may explain the differences in egg mass released at the two different depths (Wacker & Elert 2003). In addition, metabolic rates increase with temperature, so that differences in food availability at the two depths could further influence reproductive investment. Without respective to temperature differences, availability of polyunsaturated fatty acids and food quality in general resulted in changes to reproductive investment (Wacker & Elert 2003). An additional study confirms a threshold of 12 C for the onset of spawning and further identifies two spawning cohorts as the season proceeds from May to August (Borcherding 1991). Gametes are released into the water column over a period of 6 to 8 weeks, where they are fertilized and develop—30,000 to 40,000 eggs may be released by one female, though the actual number may be closer to 1.5 million. Egg release corresponds with temperature, beginning at 12C and peaking at 22C. As a result, juvenile proliferation should track these temperature changes (Hebert et al. 1988). High temperatures alongside low food availability results in reduction in the size of the gonads (Borcherding 1991).

Dreissena tends to colonize structures below 1.2 m. Zebra mussels may spread to larger structures through water intake pipes. Zebra mussels tend to increase water clarity through digestion of suspended sediments—for this reason zebra mussels have been intentionally stocked in lakes outside North America. Food deprived mussels fed indiscriminately on particles of all sizes, but satiated individuals felt only on those in a much smaller size range (MacIsaac 1996).

Ambient temperature, seston concentration, and mussel size frequency are three factors that influence Dreissena filtering impact. Maximal filtering rate has been hypothesized to be 5 and 20 C, declining outside of this range. Ingestion rate may also be temperature dependent; however, the results of experiments investigating the effect of temperature are variable. It may still be an important performance factor, and impose a range limitation in southern states. Depending on size, zooplankton may succumb to or evade ingestion by zebra mussels—smaller individuals cannot escape the inflow current and are not rejected, whereas larger individuals may dodge the current or be expelled as a result of irritating the feeding apparatus (MacIsaac 1996).






```{r, echo=FALSE}
dat <- read.csv("NAS-Data-Download.csv")
library(dplyr)
library(ggplot2)

```


```{r, echo = FALSE, eval=FALSE}

# Checking to unique HUC ID Numbers
as.data.frame(table(dat$HUC.8.Number))
# There are 350 unique HUC ID numbers
# That means there are 350 different locations for this data
```


```{r, echo = FALSE}
# This chunk will be to filter the unnecessary parts of the data
fil_dat <- select(dat, Specimen.Number, Species.ID, Group, Family, Scientific.Name, 
                  Common.Name, Latitude, Longitude, Source, Accuracy, Drainage.Name,
                  HUC.8.Number, Year, Month, Day, Status)

```

#Initial Visualization
The graph below is one way in which the data set from the USGS zebra mussel data can be visualized. The horizontal axis represents the HUC Number and it is plotted again the state in which that HUC resides. The color of the point exhibits the time when the data was collected. There appears to be an even spread of years throughout the plot given by the various color gradients. 
```{r, echo = FALSE}
# Just looking at some different ways to visualize and understand the data
ggplot(fil_dat, aes(x=dat$HUC.8.Number, y=dat$State)) + geom_point(aes(col=dat$Year), size=3) +labs(title="HUC Number vs. State", subtitle="From USGS Zebra Mussel Dataset", y="State", x="HUC Number", col = "Years") 


# ggplot(fil_dat, aes(x=dat$HUC.8.Number, y=dat$Year)) + geom_point(aes(col=dat$State), size=3) +labs(title="HUC Number vs. Year", subtitle="From USGS Zebra Mussel Dataset", y="Year", x="HUC Number")


counts <- table(fil_dat$HUC.8.Number)
barplot(counts, main="HUC Distribution", 
   xlab="Number of Occurances in HUC")

# Look into the other large points from the HUC Distribution

# max(counts) #684 This occurs in Oneida Lake in NY
# which(table(fil_dat$HUC.8.Number) == 684) #4140202
# 
# head(fil_dat[fil_dat$HUC.8.Number == 414020200,])
# 
# 
# table(fil_dat$HUC.8.Number)


```


#Maps and graph by state and year
```{r, echo=FALSE}
#dat <- read.csv("C:/Users/kmb057/Documents/FundamentalsQuantReasoning/QuantReasoning/NAS-Data-Download.csv")

#remove 0,0 coordinate row
dat <- dat[-5810,]

library(ggplot2)
library(maps)
library(gganimate)
library(dplyr)

#map of the 50 states
US <- map_data("state") 

states <- ggplot(data = US) + 
  geom_polygon(aes(x = long, y = lat, group = group), color = "white") + 
  coord_fixed(1.3) +
  guides(fill=FALSE)

#show change through time from earliest to latest
dat$Year <- sort(dat$Year,decreasing=F)

map1 <- states + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=dat,alpha=0.2)+transition_states(dat$Year)

#population in 2019
latest <- dat[dat$Year=="2019",]
map2 <- states + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=latest,alpha=0.2)

#animation of only those established or collected after 2000
established_collected_post_2000 <- filter(dat[dat$Year>=2000,]) %>%
  filter(Status=="established"|Status=="collected")

established_collected_post_2000$Year <- sort(established_collected_post_2000$Year,decreasing=F)

map3 <- states + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=established_collected_post_2000,alpha=0.2)+transition_states(established_collected_post_2000$Year)+shadow_mark()




#Plot number of zebra mussels sightings per year 
Summarize_sightings_by_year_month <- dat %>% count(Year, Month,State)
plot(Summarize_sightings_by_year_month$Year,Summarize_sightings_by_year_month$n)



ggplot(Summarize_sightings_by_year_month, aes(fill=Summarize_sightings_by_year_month$State, y=Summarize_sightings_by_year_month$n, x=Summarize_sightings_by_year_month$Year)) +labs(x = "Year",y="Occurrences",colour="State") + geom_bar(position="dodge", stat="identity")+theme(legend.position = "none")

  
#filter data for only Great Lakes lat and long
BorderingGreatLakes <- c("New York","Pennsylvania","Ohio","Indiana","Michigan","Illinois","Wisconsin","Minnesota")
PlottingBorder <- map_data("state", region=BorderingGreatLakes)

Borders <- ggplot(data = PlottingBorder) + 
  geom_polygon(aes(x = long, y = lat, group = group), color = "white") + 
  coord_fixed(1.3) +
  guides(fill=FALSE)+scale_y_continuous(breaks=seq(from=30,to=50,by=1))+scale_x_continuous(breaks=seq(from=-100,to=-70,by=1))


Superior <- dat %>% filter(Latitude>=46 & Latitude<=50) %>% filter(Longitude<=-84.5 & Longitude>=-92)
map4 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Superior,alpha=0.2)

Michigan <- dat %>% filter(Latitude>=42.5 & Latitude<=46) %>% filter(Longitude<=-84.5 & Longitude>=-88)
map5 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Michigan,alpha=0.2)

Huron <- dat %>% filter(Latitude>=43 & Latitude<=46) %>% filter(Longitude<=-81.5 & Longitude>=-84)
map6 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Huron,alpha=0.2)

Erie <- dat %>% filter(Latitude>=41.5 & Latitude<=43) %>% filter(Longitude<=-79 & Longitude>=-84)
map7 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Erie,alpha=0.2)

Ontario <- dat %>% filter(Latitude>=43.25 & Latitude<=44) %>% filter(Longitude<=-78 & Longitude>=-80)
map8 <- Borders + geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Ontario,alpha=0.2)
  
 #all great Lakes plotted
 GreatLakes <- Borders + 
  geom_point(aes(x=Longitude,y=Latitude,color=Status),data=Superior,alpha=0.2)+
  geom_point(aes(x=Longitude,y=Latitude,color="blue"),data=Michigan,alpha=0.2)+
  geom_point(aes(x=Longitude,y=Latitude,color="green"),data=Huron,alpha=0.2)+
  geom_point(aes(x=Longitude,y=Latitude,color="yellow"),data=Erie,alpha=0.2)+
  geom_point(aes(x=Longitude,y=Latitude,color="purple"),data=Ontario,alpha=0.2)
 
 GreatLakes
```
```{r}
#number of specimens found in each lake`
sup_occ <- unique(Superior$Specimen.Number)
mich_occ <- unique(Michigan$Specimen.Number)
huron_occ <- unique(Huron$Specimen.Number)
erie_occ <- unique(Erie$Specimen.Number)
ont_occ <- unique(Ontario$Specimen.Number)
zm_occ <- c(sup_occ, mich_occ, huron_occ, erie_occ, ont_occ)

SupOcc <- nrow(table(sup_occ))
MichOcc <- nrow(table(mich_occ))
HurOcc <- nrow(table(huron_occ))
EriOcc <- nrow(table(erie_occ))
OntOcc <- nrow(table(ont_occ))
zm_occ <- c(SupOcc, MichOcc, HurOcc, EriOcc, OntOcc)
```
One occurrence indicates one collection effort--anywhere from one to thousands of zebra mussels could be included. Lake Michigan has the highest number of occurrences at 610, while Lake Superior has the least, at 36.
```{r}

CountYear <- lm(n~Year, Summarize_sightings_by_year_month)
summary(CountYear)
ggplot(Summarize_sightings_by_year_month, aes(x=Summarize_sightings_by_year_month$Year, y=Summarize_sightings_by_year_month$n)) + geom_point() + geom_smooth(method = "lm") + labs(title="Occurrences By Year", y="Count",x="Year")

```
There is no relationship between number of occurrences and year. Number of sightings does not increase or decrease as time goes on.

#Linear Regression of Size of Lake and HUC Counts
```{r SizeLR, echo=FALSE}
#Want to use this chunk to see if there is a correlation between number of occurances in each lake and their size. Linear Regression Analysis. If there isn't a positive correlation then would it be more around tourist areas? or more populated with boating areas. 
 
 # https://www.britannica.com/place/Great-Lakes
 
 #Superior, Michigan, Huron, Erie, Ontario sizes in square miles
 size_of_lake <- c(31700, 22300, 23000, 9910, 7340)
 
 
 sup_counts <-nrow(table(Superior$HUC.8.Number))
 mich_counts <-nrow(table(Michigan$HUC.8.Number))
 huron_counts <-nrow(table(Huron$HUC.8.Number))
 erie_counts <- nrow(table(Erie$HUC.8.Number))
 ont_counts <- nrow(table(Ontario$HUC.8.Number))
 zm_HUC_occurances <-c(sup_counts, mich_counts, huron_counts, erie_counts, ont_counts)
 
 sizeDat <- data.frame(size_of_lake, zm_HUC_occurances)

 
 HUCmodelSize <- lm(zm_HUC_occurances~size_of_lake, sizeDat)
 summary(HUCmodelSize)
 sizeCoef <- coef(HUCmodelSize)[2]
 
 
# Call:
#lm(formula = zm_occurances ~ size_of_lake, data = sizeDat)

#Residuals:
#     1      2      3      4      5 
#-6.915 14.778 -4.348  6.011 -9.526 

#Coefficients:
#              Estimate Std. Error t value Pr(>|t|)
#(Intercept)  1.020e+01  1.214e+01    0.84    0.462
#size_of_lake 1.802e-04  5.811e-04    0.31    0.777
 
 # The p-value associated with the coefficient is not a significant value, and therefore not statistically significant. From the model, its states that a unit increase in the size of the lake leads to a 0.0001802 increase in the HUC Occurances.  

#Residual standard error: 11.72 on 3 degrees of freedom
#Multiple R-squared:  0.03104,	Adjusted R-squared:  -0.2919 
#F-statistic: 0.0961 on 1 and 3 DF,  p-value: 0.7768
 
 
 ggplot(sizeDat, aes(x=size_of_lake, y=zm_HUC_occurances)) + geom_point() + geom_smooth(method = "lm") + labs(title="Size of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset", y="HUC Occurrences", x="Size of Lake") 
 
 # There is a confidence band around the regression line as well. 

#number of specimens found in each lake`
sup_occ <- unique(Superior$Specimen.Number)
mich_occ <- unique(Michigan$Specimen.Number)
huron_occ <- unique(Huron$Specimen.Number)
erie_occ <- unique(Erie$Specimen.Number)
ont_occ <- unique(Ontario$Specimen.Number)
zm_occ <- c(sup_occ, mich_occ, huron_occ, erie_occ, ont_occ)

SupOcc <- nrow(table(sup_occ))
MichOcc <- nrow(table(mich_occ))
HurOcc <- nrow(table(huron_occ))
EriOcc <- nrow(table(erie_occ))
OntOcc <- nrow(table(ont_occ))
zm_occ <- c(SupOcc, MichOcc, HurOcc, EriOcc, OntOcc)
 
```
A linear regression analysis was conducted after collecting data on the size of the lake and filtering out the data to only look at the HUC zebra mussel occurrences in the areas of the Great Lakes. The reason for this analysis is to understand if there is a relationship between the size of the lake in water volume, and the number of occurrences on certain HUCs. 

Once the analysis was conducted, the p-value associated with the coefficient is not a significant value, and therefore not statistically significant. From the model, its states that a unit increase in the size of the lake leads to a `r sizeCoef` increase in the HUC Occurrences. Therefore, there is not a significant relationship between the HUC Occurrences, and the size of each Great Lake. The plot also depicts the trend line with the slope of `r sizeCoef` and there are confidences bands around the trend line also. This is a small data set with only five data points though, given there are only five Great Lakes. 



```{r separateSizeCode, echo = FALSE}
# This chunk will change the years that the HUC data is being looked at in 8 year chunks
 SuperiorFirst8 <- dat %>% 
   filter(Latitude>=46 & Latitude<=50) %>% 
   filter(Longitude<=-84.5 & Longitude>=-92) %>% 
   filter(Year <=1999 & Year >= 1991)
 
 SuperiorSecond8 <- dat %>% 
   filter(Latitude>=46 & Latitude<=50) %>% 
   filter(Longitude<=-84.5 & Longitude>=-92) %>% 
   filter(Year <= 2008 & Year >= 2000)

 SuperiorThird8 <- dat %>% 
   filter(Latitude>=46 & Latitude<=50) %>% 
   filter(Longitude<=-84.5 & Longitude>=-92) %>% 
   filter(Year <= 2015 & Year >= 2009)
 
MichiganFirst8 <- dat %>% 
  filter(Latitude>=42.5 & Latitude<=46) %>% 
  filter(Longitude<=-84.5 & Longitude>=-88) %>% 
   filter(Year <=1999 & Year >= 1991)

MichiganSecond8 <- dat %>% 
  filter(Latitude>=42.5 & Latitude<=46) %>% 
  filter(Longitude<=-84.5 & Longitude>=-88) %>% 
   filter(Year <=2008 & Year >= 2000)

MichiganThird8 <- dat %>% 
  filter(Latitude>=42.5 & Latitude<=46) %>% 
  filter(Longitude<=-84.5 & Longitude>=-88) %>% 
   filter(Year <=2015 & Year >= 2009)

HuronFirst8 <- dat %>% 
  filter(Latitude>=43 & Latitude<=46) %>% 
  filter(Longitude<=-81.5 & Longitude>=-84) %>% 
   filter(Year <=1999 & Year >= 1991)

HuronSecond8 <- dat %>% 
  filter(Latitude>=43 & Latitude<=46) %>% 
  filter(Longitude<=-81.5 & Longitude>=-84) %>% 
   filter(Year <=2008 & Year >= 2000)

HuronThird8 <- dat %>% 
  filter(Latitude>=43 & Latitude<=46) %>% 
  filter(Longitude<=-81.5 & Longitude>=-84) %>% 
   filter(Year <=2015 & Year >= 2009)

ErieFirst8 <- dat %>% 
  filter(Latitude>=41.5 & Latitude<=43) %>%
  filter(Longitude<=-79 & Longitude>=-84) %>% 
   filter(Year <=1999 & Year >= 1991)

ErieSecond8 <- dat %>% 
  filter(Latitude>=41.5 & Latitude<=43) %>%
  filter(Longitude<=-79 & Longitude>=-84) %>% 
   filter(Year <=2008 & Year >= 2000)

ErieThird8 <- dat %>% 
  filter(Latitude>=41.5 & Latitude<=43) %>%
  filter(Longitude<=-79 & Longitude>=-84) %>% 
   filter(Year <=2015 & Year >= 2009)

OntarioFirst8 <- dat %>% 
  filter(Latitude>=43.25 & Latitude<=44) %>% 
  filter(Longitude<=-78 & Longitude>=-80) %>% 
   filter(Year <=1999 & Year >= 1991)

OntarioSecond8 <- dat %>% 
  filter(Latitude>=43.25 & Latitude<=44) %>% 
  filter(Longitude<=-78 & Longitude>=-80) %>% 
   filter(Year <=2008 & Year >= 2000)

OntarioThird8 <- dat %>% 
  filter(Latitude>=43.25 & Latitude<=44) %>% 
  filter(Longitude<=-78 & Longitude>=-80) %>% 
   filter(Year <=2015 & Year >= 2009)

```

To conduct further analysis, that data used used the time span of the years 1991-2015 and the data was then split into three different different time period. The time periods are [1991-1999], [2000-2008], and [2009-2015]. Each time period has a linear regression analysis associated with it and the details for each will be described along with a graph of each trend line. 



```{r regressFirst8, echo = FALSE}
  ## This is the first 8 years of the data (1991-1999)

sup_counts_F8 <-nrow(table(SuperiorFirst8$HUC.8.Number))
 mich_counts_F8 <-nrow(table(MichiganFirst8$HUC.8.Number))
 huron_counts_F8<-nrow(table(HuronFirst8$HUC.8.Number))
 erie_counts_F8 <- nrow(table(ErieFirst8$HUC.8.Number))
 ont_counts_F8 <- nrow(table(OntarioFirst8$HUC.8.Number))
 zm_HUC_occurances_F8 <-c(sup_counts_F8, mich_counts_F8, huron_counts_F8, erie_counts_F8, ont_counts_F8)

 HUCmodelSizeF8 <- lm(zm_HUC_occurances_F8~size_of_lake, sizeDat)
 summary(HUCmodelSizeF8)
 sizeCoefF8 <- coef(HUCmodelSizeF8)[2]
#  
# Call:
# lm(formula = zm_HUC_occurances_F8 ~ size_of_lake, data = sizeDat)
# 
# Residuals:
#        1        2        3        4        5 
# -5.83035  9.09380 -0.04949  2.62999 -5.84395 
# 
# Coefficients:
#               Estimate Std. Error t value Pr(>|t|)
# (Intercept)  5.3414746  7.5144357   0.711    0.528
# size_of_lake 0.0002047  0.0003596   0.569    0.609
# 
# Residual standard error: 7.252 on 3 degrees of freedom
# Multiple R-squared:  0.09748,	Adjusted R-squared:  -0.2034 
# F-statistic: 0.324 on 1 and 3 DF,  p-value: 0.609

 
 sizeDatF8 <- data.frame(size_of_lake, zm_HUC_occurances_F8)

 
 ggplot(sizeDatF8, aes(x=size_of_lake, y=zm_HUC_occurances_F8)) + geom_point() + geom_smooth(method = "lm") + labs(title="Size of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset For First 8 Years", y="HUC Occurrences", x="Size of Lake") 
 
```

The p-value associated with the coefficient from the data in the range of 1991-1999 is not a significant value, and therefore not statistically significant. From the model, its states that a unit increase in the size of the lake leads to a `r sizeCoefF8` increase in the HUC Occurrences between the years of 1991-1999. Therefore, there is not a significant relationship between the HUC Occurrences, and the size of each Great Lake during that time frame. The plot also depicts the trend line with the slope of `r sizeCoefF8` and there are confidences bands around the trend line also. 

```{r regressSecond8, echo = FALSE}
  ## This is the second 8 years of the data (2000-2008)

sup_counts_S8 <-nrow(table(SuperiorSecond8$HUC.8.Number))
 mich_counts_S8 <-nrow(table(MichiganSecond8$HUC.8.Number))
 huron_counts_S8<-nrow(table(HuronSecond8$HUC.8.Number))
 erie_counts_S8 <- nrow(table(ErieSecond8$HUC.8.Number))
 ont_counts_S8 <- nrow(table(OntarioSecond8$HUC.8.Number))
 zm_HUC_occurances_S8 <-c(sup_counts_S8, mich_counts_S8, huron_counts_S8, erie_counts_S8, ont_counts_S8)

 HUCmodelSizeS8 <- lm(zm_HUC_occurances_S8~size_of_lake, sizeDat)
 summary(HUCmodelSizeS8)
 sizeCoefS8 <- coef(HUCmodelSizeS8)[2]
 
#  Call:
# lm(formula = zm_HUC_occurances_S8 ~ size_of_lake, data = sizeDat)
# 
# Residuals:
#       1       2       3       4       5 
# -1.8982  1.7347  0.7621  3.2509 -3.8495 
# 
# Coefficients:
#                Estimate Std. Error t value Pr(>|t|)
# (Intercept)   6.136e+00  3.415e+00   1.797    0.170
# size_of_lake -3.905e-05  1.634e-04  -0.239    0.827
# 
# Residual standard error: 3.295 on 3 degrees of freedom
# Multiple R-squared:  0.01868,	Adjusted R-squared:  -0.3084 
# F-statistic: 0.0571 on 1 and 3 DF,  p-value: 0.8265
 
 sizeDatS8 <- data.frame(size_of_lake, zm_HUC_occurances_S8)

 
 ggplot(sizeDatS8, aes(x=size_of_lake, y=zm_HUC_occurances_S8)) + geom_point() + geom_smooth(method = "lm") + labs(title="Size of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset For Second 8 Years", y="HUC Occurrences", x="Size of Lake") 
 
```

The p-value associated with the coefficient from the data in the range of 2000-2008 is not a significant value, and therefore not statistically significant. From the model, its states that a unit increase in the size of the lake leads to a `r sizeCoefS8` increase in the HUC Occurrences between the years of 2000-2008. Therefore, there is not a significant relationship between the HUC Occurrences, and the size of each Great Lake during that time frame. The plot also depicts the trend line with the slope of `r sizeCoefS8` and there are confidences bands around the trend line also. This relationship is also negative compared the the first 8 years, which had a positive relationship. Even though neither of the slopes were significant, there could be a change occurring in those time periods that need to be explored. 

```{r regressThird8, echo = FALSE}
  ## This is the third 8 years of the data (2009-2015)

sup_counts_T8 <-nrow(table(SuperiorThird8$HUC.8.Number))
 mich_counts_T8 <-nrow(table(MichiganThird8$HUC.8.Number))
 huron_counts_T8<-nrow(table(HuronThird8$HUC.8.Number))
 erie_counts_T8 <- nrow(table(ErieThird8$HUC.8.Number))
 ont_counts_T8 <- nrow(table(OntarioThird8$HUC.8.Number))
 zm_HUC_occurances_T8 <-c(sup_counts_T8, mich_counts_T8, huron_counts_T8, erie_counts_T8, ont_counts_T8)

 HUCmodelSizeT8 <- lm(zm_HUC_occurances_T8~size_of_lake, sizeDat)
 summary(HUCmodelSizeT8)
 sizeCoefT8 <- coef(HUCmodelSizeT8)[2]
 sizeDatT8 <- data.frame(size_of_lake, zm_HUC_occurances_T8)
# 
#  Call:
# lm(formula = zm_HUC_occurances_T8 ~ size_of_lake, data = sizeDat)
# 
# Residuals:
#       1       2       3       4       5 
#  0.2118  1.4717 -2.4732  4.4961 -3.7063 
# 
# Coefficients:
#                Estimate Std. Error t value Pr(>|t|)
# (Intercept)   4.284e+00  3.890e+00   1.101    0.351
# size_of_lake -7.874e-05  1.862e-04  -0.423    0.701
# 
# Residual standard error: 3.754 on 3 degrees of freedom
# Multiple R-squared:  0.05628,	Adjusted R-squared:  -0.2583 
# F-statistic: 0.1789 on 1 and 3 DF,  p-value: 0.7008

 
 
 ggplot(sizeDatT8, aes(x=size_of_lake, y=zm_HUC_occurances_T8)) + geom_point() + geom_smooth(method = "lm") + labs(title="Size of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset For Third 8 Years", y="HUC Occurrences", x="Size of Lake") 
 
```

The p-value associated with the coefficient from the data in the range of 2009-2015 is not a significant value, and therefore not statistically significant. From the model, its states that a unit increase in the size of the lake leads to a `r sizeCoefT8` increase in the HUC Occurrences between the years of 2009-2015. Therefore, there is not a significant relationship between the HUC Occurrences, and the size of each Great Lake during that time frame. The plot also depicts the trend line with the slope of `r sizeCoefT8` and there are confidences bands around the trend line also. This coefficient is negative and larger than the previous 8 year period. However, the overall regression analyis shows a positive non significant trend, which insinuates that the first 8 year period has a large affect on the regression analysis since it is the only positive trend out of the three time periods. 


```{r Temp, echo = FALSE}
##  Temperature data
library(readxl)
lakes_92_18 <- read_excel("great lakes avg temp 1992-2018.xlsx")
every_lake <- read_excel("every lake 1995-2018.xlsx")

#This graph shows the average temperatures for the lakes over the course of the year
#IT'S A GOOD GRAPH

plot(lakes_92_18$Day,lakes_92_18$Superior, col= "blue", xlim= c(1,365), ylim= c(0,30), 
     xlab= "Day of the Year", ylab= "Temperature (Celsius)", main= "Average Temp 1992-2018")
points(lakes_92_18$Day,lakes_92_18$Michigan, col= "orange")
points(lakes_92_18$Day,lakes_92_18$Huron, col="green" )
points(lakes_92_18$Day,lakes_92_18$Erie, col= "red")
points(lakes_92_18$Day,lakes_92_18$Ontario, col= "purple")
legend(0, 27, legend=c("Superior","Michigan","Huron","Erie","Ontario"), 
       col=c("blue","orange","green","red","purple"), lty=1, title= "Lake")
# #This shows the consistency of the temperature of Lake Michigan over the years
Mich_Temp <- every_lake%>%
select(Year,Day,Mich.)
# 
#   plot(Temp)
# 
ggplot(Mich_Temp, aes(x= Year, y= Mich.))+geom_point()


#This shows the temperature of Lake Superior 
sup_temp <-  every_lake%>%
  select(Year,Day,Sup.)
ggplot(sup_temp, aes(x= Year, y= Sup.))+geom_point()


```

The average temperature of each lake throughout the year shows that Lake Erie get sthe warmest and Lake Superior is the coldest. When specifically looked at, Lake Superior has a high temperature that is about 5 degrees lower than the high temperature of Lake Michigan. There is a lower density of zebra mussels in Lake Superior due to its lower temperatures and harsher conditions.  

```{r TempLR, echo = FALSE}


superiorAvgTemp <- sum(lakes_92_18$Superior)/366
michiganAvgTemp <- sum(lakes_92_18$Michigan)/366
huronAvgTemp <- sum(lakes_92_18$Huron)/366
erieAvgTemp <- sum(lakes_92_18$Erie)/366
ontarioAvgTemp <- sum(lakes_92_18$Ontario)/366

avgTemp <-c(superiorAvgTemp, michiganAvgTemp, huronAvgTemp,erieAvgTemp,ontarioAvgTemp)


 sup_counts <-nrow(table(Superior$HUC.8.Number))
 mich_counts <-nrow(table(Michigan$HUC.8.Number))
 huron_counts <-nrow(table(Huron$HUC.8.Number))
 erie_counts <- nrow(table(Erie$HUC.8.Number))
 ont_counts <- nrow(table(Ontario$HUC.8.Number))
 zm_HUC_occurances <-c(sup_counts, mich_counts, huron_counts, erie_counts,  
                       ont_counts)
 
 TempDat <- data.frame(avgTemp, zm_HUC_occurances)

 
 HUCmodelTemp <- lm(zm_HUC_occurances~avgTemp, TempDat)
 summary(HUCmodelTemp)
 
#  Call:
# lm(formula = zm_HUC_occurances ~ avgTemp, data = TempDat)
# 
# Residuals:
#        1        2        3        4        5 
#  -0.6765  14.9544  -2.9716   1.5067 -12.8131 
# 
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)
# (Intercept)    0.867     29.363    0.03    0.978
# avgTemp        1.388      3.150    0.44    0.689
# 
# Residual standard error: 11.54 on 3 degrees of freedom
# Multiple R-squared:  0.06075,	Adjusted R-squared:  -0.2523 
# F-statistic: 0.194 on 1 and 3 DF,  p-value: 0.6894

 
 ggplot(sizeDatT8, aes(x=avgTemp, y=zm_HUC_occurances_T8)) + geom_point() + geom_smooth(method = "lm") + labs(title="Temperature of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset", y="HUC Occurrences", x="Temperature of Lake") 

```

Analysis to be done on this, not significant values here.

```{r, echo=FALSE}
 # Superior, Michigan, Huron, Erie, Ontario sizes in cubic miles
volume_of_lake <- c(2900, 1180, 850, 116, 393)


sup_counts <-nrow(table(Superior$HUC.8.Number))
mich_counts <-nrow(table(Michigan$HUC.8.Number))
huron_counts <-nrow(table(Huron$HUC.8.Number))
erie_counts <- nrow(table(Erie$HUC.8.Number))
ont_counts <- nrow(table(Ontario$HUC.8.Number))
zm_HUC_occurances <-c(sup_counts, mich_counts, huron_counts, erie_counts, ont_counts)

volumeDat <- data.frame(volume_of_lake, zm_HUC_occurances)


HUCmodelVolume <- lm(zm_HUC_occurances~volume_of_lake, volumeDat)
summary(HUCmodelVolume)




ggplot(volumeDat, aes(x=volume_of_lake, y=zm_HUC_occurances)) + geom_point() + geom_smooth(method = "lm") + labs(title="Volume of Lake vs. HUC Occurrences", subtitle="From USGS Zebra Mussel Dataset", y="HUC Occurrences", x="Volume of Lake") 


```

